---
title: "Bootstrapping Confidence Intervals"
output: html_notebook
---

## Introduction


#### Understanding confidence intervals

Here is a quick note on why we use bootstrapping and some different approaches.

*Reference: Notes <a href="http://users.stat.umn.edu/~helwig/notes/bootci-Notes.pdf">here</a> and <a href="https://statweb.stanford.edu/~ckirby/brad/papers/2018Automatic-Construction-BCIs.pdf">here</a>*

We can use bootstrapping to 'brute force' the confidence intervals of some statistics that might be hard to derive analytically (for example, median or r2 from regression).

*Be careful of the interpretation of confidence intervals though, it is often stated incorrectly. A 99% confidence interval does not mean that there is a 99% probability the true value lies within this range: it either lies within the range, or it doesn't, there is no likelihood involved as this is a frequenist definition. What a confidence interval does mean is that if we repeated the proceedure many, many times we would expect the intervals to contain the true value of $\theta$ for 99% of the proceedures. See more details <a href="https://chris-kelly.github.io/ds_blog/bayesian/bayesian_v2.nb.html">here.</a>*

This interpretation fits well with bootstrapping, where we essentially re-run the proceedure many times (just on the existing sample, rather than on the population).

#### Describing confidence intervals

A symmetric 100(1-$\alpha$)% confidence interval has the form  $\hat{\theta} \pm t_{\alpha/2}\sigma_{\hat{\theta}}$.

(where $\hat{\theta}$ is the estimate of $\hat{\theta}$, $\sigma_{\hat{\theta}}$ is the standard error of $\hat{\theta}$).

We thus might define two proporties to describe a confidence interval:

* Length: $\hat{\theta}_{high} - \hat{\theta}_{low}$
* Shape: $\frac{\hat{\theta}_{high} - \hat{\theta}}{\hat{\theta} - \hat{\theta}_{low}}$ (where shape > 1 indicates positive skew, <  is negative skew)

*Note: A narrower CI is not necesarily a good CI - length and shape only important is coverage probabilities are accurate!*


#### Defining coverage probability accuracy:

We can define coverage probability as the proportion of the time that the interval contains the true value of interest. The nominal coverage probability is often set at 95%, but the actual coverage probability could be more or less if any assumptions are not met. The lower the sample size $n$, the more likely the actual coverage probability differs from the nominal coverage probability.

We may then determine a method for defining confidence intervals is first-order accurate if the non-coverage probability on each side differs from the nominal value by $O(n^{-1/2})$. 

We determine the method is second-order accurate if the non-coverage probability on each side differs from the nominal value by $O(n^{-1})$.

It is thus desirable to find methods that have better coverage probability!

## Classic standard error for assuming confidence intervals:

We estimate this using the classic formula from the sample:

$$
\hat{\theta} \pm t_{\alpha/2}\hat{\sigma}_{\hat{\theta}}
$$

```{r classic_conf}
set.seed(1)
x = rnorm(50,mean=1)
n <- length(x)

SE = sd(x)/sqrt(n) # Est standard deviation of sample means
student_t_lookup <- qt(0.975,df=n-1) 

conf_intervals_classic <- c(mean(x) - qt(0.975,df=n-1)*SE, mean(x) + qt(0.975,df=n-1)*SE)
conf_intervals_classic
```

## Bootstrap SE:

Method: 
Sample with replacement, calculate $\sigma$ each time, use mean in calc


$$
\hat{\theta} \pm {t_{\alpha/2}\hat{\sigma}_{\hat{B}}}
$$

* No advantage of classic standard error tbh
* Tends to be too narrow for small n
* Is poor if distribution highly skewed
* Only first order accurate

```{r bootstrap_se}
bsamp <- sapply(1:10^5, FUN= function(i) {sample(x,replace=T)})
SE_b = sd(apply(bsamp,2,mean)) # Standard deviation of sample means
conf_intervals_bootstrap_se <- c(mean(x) - qt(0.975,df=n-1)*SE_b, mean(x) + qt(0.975,df=n-1)*SE_b)
conf_intervals_bootstrap_se
```

## Bootstrap percentiles:

Method: 
Sample with replacement $B$ times, calculate $\hat{\theta}$ each time and put in order of size.

90% confidence interval lies in:

$$
[\hat{\theta}_{0.05 \times B},\hat{\theta}_{0.95 \times B}]
$$

* First order accurate (not second order)
* Range preserving, transformation invariant.
* Can deal with skewness of distribution
* Too narrow if small n (even worse than Bootstrap SE! - like using $z_{\alpha/2}\hat{\sigma}_{\hat{\theta}}$ instead of $t_{\alpha/2}\hat{\sigma}_{\hat{\theta}}$, see "Expanded percentile confidence intervals")
* Only first order accurate

```{r bootstrap_percentile}
bootstrap_percentile <- quantile(apply(bsamp,2,mean),c(0.025,0.975),names = F)
bootstrap_percentile
```

## Expanded percentile confidence interval:

Percentile CI is comparable to using Z distribution (i.e. $z_{\alpha/2}\frac{\hat{\sigma}}{\sqrt{n}}$) instead of t distribution (i.e. $t_{\alpha/2}\frac{\hat{s}}{\sqrt{n}}$).

*Note $s$ is the sample standard deviation (with $n-1$ degrees of freedom), as opposed to the population standard deviation (where we just divide by $n$).*

Hence we can correct this by a factor $a_{\alpha,n}=\frac{t_{\alpha/2}}{z_{\alpha/2}}\times\frac{s}{\hat{\sigma}}$, i.e.

$$
z_{\alpha/2}\frac{\hat{\sigma}}{\sqrt{n}} \times \left( \frac{t_{\alpha/2}}{z_{\alpha/2}}\times\frac{s}{\hat{\sigma}} \right) = t_{\alpha/2}\frac{\hat{s}}{\sqrt{n}}
$$

In practice though, we don't apply this correction to both sides of the interval (as would no longer be transformation invariant) but to the quantiles of the bootstrap distribution. 
Hence we need to find a new alpha for Z distribution, $\alpha'$, such that it equals the t-distribution estimation:

$$
\hat{\theta} + z_{\alpha'/2} \frac{\hat{\sigma}}{\sqrt{n}} =\hat{\theta} + t_{\alpha/2} \frac{s}{\sqrt{n}} \\
\Rightarrow z_{\alpha'/2} = t_{\alpha/2} \times \frac{s}{\hat{\sigma}} = t_{\alpha/2} \times \sqrt{\frac{n}{n-1}} \\
\Rightarrow \alpha/2 = \phi \left(\sqrt{\frac{n}{n-1}} \right)
$$

We can then use this alpha instead to determine which percentiles to pick.

```{r expanded_percentile}
alpha2 <-  pnorm(sqrt(n/(n-1))*qt(.025,df=n-1))
expanded_bootstrap_percentile <- quantile(apply(bsamp,2,mean),c(alpha2,1-alpha2))
expanded_bootstrap_percentile
```

* Corrects for narrowness bias of percentile CIs
* Range preserving and transformation invariant
* No correction for bias, and doesn’t fully correct for skewness
* Only first order accurate

## Bias corrected and Accerlated bootstrap CIs (BCa):

* Have intervals of the form $[\hat{\theta_{\alpha_{low}}},\hat{\theta_{\alpha_{high}}}]$ but these do not necessarily map to 100$\alpha$-th and 100 $(1-\alpha)$-th percentiles 
* Depends on acceleration parameter $\hat{a}$ and bias-correction factor $\hat{z_0}$


$$
\alpha_{low} = \Phi \left(\hat{z_0} + \frac{\hat{z_0} + z_{(\alpha)}}{1-\hat{a}(\hat{z_0} + z_{(\alpha)})} \right) \\
\alpha_{high} = \Phi \left(\hat{z_0} + \frac{\hat{z_0} + z_{(1-\alpha)}}{1-\hat{a}(\hat{z_0} + z_{(1-\alpha)})} \right) \\
$$

Where $\Phi(.)$ is the CDF of the standard normal and $z_{(\alpha)}$ is the 100$\alpha$-th percentile of the standard normal

Note then, if $\hat{a} = \hat{z_0} = 0$, then $\alpha_1 = \alpha$; and $\alpha_2 = 1 - \alpha;$ (same as percentile interval)

The bias correction factor measures the median bias of $\hat{\theta}$ (i.e. the difference between the bootstrapped median $\hat{\theta_b}$ and the sample mean $\hat{\theta}$)

The acceleration factor is calculated using a jackknife approach (where jacknife is the average of the mean $\hat{\theta_i}$ when leaving one sample out at a time, e.g. 

$$
\hat{\theta }_{i} = \frac{1}{n-1} \sum_{j \neq i}^{n}{\theta_i} \\
\hat{\theta }_{(.)} = \frac{1}{n}\sum_{i=1}^{n}{\hat{\theta }_{i}} \\
\hat{a} = \frac{ \sum_{i=1}^{n}{\left( \hat{\theta }_{(.)} - \hat{\theta }_{(i)} \right)^3}}{6 \left[\sum_{i=1}^{n}{\left( \hat{\theta }_{(.)} - \hat{\theta }_{(i)} \right)^2} \right]^{3/2}}
$$


* Range preserving and transformation invariant, works well for a variety of parameters
* Second order accurate as well!
* Less intuitive, requires estimaton of acceleration and bias-correction


```{r}
alpha = 0.025
n <- length(x)
nboot = 10^5

z0 <- qnorm(sum(apply(bsamp,2,mean)<mean(x))/dim(bsamp)[2]) #  bias_correction_factor
jackknife_i <- sapply(1:length(x),FUN = function(i) {mean(x[-i])}) # mean of 'leave on out'
a_hat <- sum((mean(jackknife_i) - jackknife_i)^3) / 6*(sum(mean(jackknife_i) - jackknife_i)^2)^(3/2) # acceleration_factor
alpha_lo <- pnorm(z0+(z0+qnorm(alpha))/(1-a_hat*(z0+qnorm(alpha)))) # low alpha
alpha_hi <- pnorm(z0+(z0+qnorm(1-alpha))/(1-a_hat*(z0+qnorm(1-alpha)))) # high alpha

bca_percentile <- quantile(apply(bsamp,2,mean),probs=c(alpha_lo,alpha_hi),names = F)
bca_percentile

# Note there is a function in bootstrap to do this: "bcanon"
```


